Cover letter for the original IJRR submission

### 2017-10-02 reviews_tro/reviewer_response.tex

\quot{*I think I understand what the authors mean by solving at
"interactive" rates, but this is worth defining. In my mind, it means
something like "before the human operator gets bored", or roughly > 0.5
Hz. But since the authors make the claim that this is the "first
interactive implementation of a contact planner", then that statement
should be made quantitative and verifiable}
\as{We agree with the relevant comment, and updated the paper accordingly.}
\qt{(End of Section I.A.) We define an interactive planner as one that outputs a solution in a few seconds or less.}\done

\quot{* In Section I.A. the authors refer to Bretl's work on multi-contact
planning as the inspiration for the P1,P2,P3 decomposition of the
problem. But it is not, as the authors state, always necessary to
decompose motion planning into choosing a series of quasi-static
configurations. There are a variety of feasible motions with no static
equilibrium configurations, so it is worth mentioning that this
decomposition only solves a subset of motion planning problem.}
\as{We apologize if this was not made clear enough in our first submission. In the conclusion, we indicate that the planner is limited to specific scenarios (where quasi-static motion
is always an option) ``\textit{Our method applies to any scenario where at least one
contact friction cone contains the direction opposed to the
gravity (i.e. quasi-flat)[\dots]}''. We updated the state of the art to indicate that the decomposition in general is restricitve.}
\qt{(End of before last paragraph of Section I.A) A limitation with
this kind of approach, shared by our own method, is that most of the existing planners only address a subset of the problem, because their ability to find a solution depends
on the existence of quasi-static equilibrium configurations along a feasible path, which is too restrictive in the general case.}\done

\quot{* Also in Sect. I.A, the authors state that "optimization-based
approaches only converge locally". This is not necessarily true, as
convex or mixed-integer convex optimizations can (sometimes) be solved
to a verifiable global optimum. I would suggest removing this sentence
entirely.}
\as{Thanks for the suggestion. Our phrasing was indeed misleading and inaccurate. Our intention was to explain that the geometry of the scene makes the problem highly non-convex, and
that approaches such as the CIO, if not initialized with an appropriate initial guess, might never get out of the local minima. We followed the reviewer's advice and simply removed the sentence. }\done

\quot{* I am not sure why the $C_{reach}$ hypothesis is described as the ``strong''
hypothesis in, e.g., Sect. IV. Is there a corresponding weak
hypothesis?}
\as{We apologize for the confusion, that comes from an inaccurate use of English. What we meant by ``strong'' is that the hypothesis is constraining, because
it defines a solution space different than the real one, and that we need to demonstrate that it makes sense.  We removed ``strong'' and replaced it by ``main'' to clarify. }\done

\quot{* In Sect. IV.B, the authors state that the ideal shape $B^*$ ``has no
explicit definition''. For a particular configuration  $q_0$, $B^*$ is simply
the surface of the robot, right? So I assume that this claim refers to
there being no explicit definition of B* which is the volume around
$B^*(q_0)$ for every $q_0$? I think I agree with the authors, but I found
this claim hard to follow and would appreciate some clarification}
\as{We updated the Section IV B. to better explain our claim. Indeed, as the reviewer states $B^*$  is the volume for every $q_0$. If such a volume existed, we would know exactly whether or not a given root configuration can result in a valid contact configuration.
We of course doubt that $B^*$ actually exists. }
\qt{The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
To our knowledge  $B^*$ has no explicit definition.
Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.}\done

@@

\quot{* In Sect. V.A., the authors mention that the ability to plan a contact
break and creation with the same limb allows the creation of dynamic
motions. This is quite an important ability for an otherwise
quasi-static planner, but this ability isn't described again until the
appendix. It would be worth explaining why this is true in the paper
itself. }
\as{Thanks for the suggestion. We added a paragraph to better explain this in Section V.A.}\ndone

\quot{In Sect. V.B, the authors mention inserting an additional state
between contact configurations if two or more contacts cannot be
maintained. Shouldn't this step appear somewhere in Alg. 1?}
\as{We apologize for the lack of clarity. This is actually handled by the method \textsc{RepositionContacts}. We now realize that the name is misleading. Indeed,
\textsc{RepositionContacts} either adds a new contact, or tries to reposition one already existing. In Appendix D, we state:
\textit{Otherwise, the method \textsc{RepositionContacts} is called. It repositions one end effector (either a free
limb, or the oldest active contact) towards a new contact
position if possible. }

We changed the name to \textsc{IntermediateContactState}, with the hope that it makes things clearer.}\done

\quot{* In Appendix A.A, it is mentioned that step 2 in the workspace
computation is conservative, with reference to figure 17-right. But
Fig. 17-right is the simplification of the convex hull, corresponding
to step 3, and it's actually hard to tell from the figure that it is
conservative. Not knowing anything about Blender's decimate tool, I
assume it is operating by removing vertices, which would indeed be
conservative, but some clarification would be useful here.}
\as{Indeed, the decimate tools removes vertices, and this was not explained clearly in the paper. We updated the text accordingly}
\qt{The resulting polytope can contain a very large number of faces. A last step is thus to simplify it with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.}\done

\quot{* Finally, I would suggest providing explicit git SHAs or tags for all
of the github links. I've personally made the mistake of linking to a
github project in a paper, only to have that project change and
eventually delete the specific code I was citing. }
\as{The reviewer is absolutely right. We indeed encounter the issue with other researchers that are currently using our planner. 
To address this issue, the link for the code now points at a website, where instructions are given to install the correct version of the project, on a separate branch
dedicated to the installation of the planner.
Regarding the link to the python code for the OCP that addresses P3, we provided an explicit git SHAs, as suggested by the reviewer.}
\qt{(Section VII) Both HPP and our planner can be simply downloaded and compiled by following the instructions on
\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}. \\
(Appendix C) To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software. }
\done

### 2017-10-02 body/appendix_algo.tex
Review paper according to reviewers suggestions
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

\subsubsection{Pseudo-code of the Algorithm}
\label{app:contact}

@@

\begin{algorithm}
\caption{Discretization of a path} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
		\State $list<$State$>$ $states = [s0]$
		\State $nb\_fail = 0$ 
		\State $i = 1;$ /*Current index in the list*/
		\While {$i < length(\mathbf{Q}^0)$}
			\State State $pState = last\_element(states)$
			\State State $s =$ \textsc{GenFullBody}$(pState, \mathbf{Q}^0[i])$
			\If {$s != NULL $}
				\State $nb\_fail = 0$
				\State $i += 1$
				\State \textbf{return} $\mathbf{q}^{0}$
			\Else
				\State $nb\_fail += 1$
				\If {$nb\_fail == MAX\_TRIES$}
					\State \textbf{return} $FAILURE$
				\EndIf				
				\State $s = $\textsc{RepositionContacts}$(pState)$
			\EndIf
			\State $push\_back(states, s)$
		\EndWhile
		\State \textbf{return} $states$
	\EndFunction
\end{algorithmic}
\label{alg:interpolate}
\end{algorithm}

At each step, \textsc{GenFullBody} is called with the previous state as a parameter, as well
as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
in computing a configuration with only one contact switch occurring.
Otherwise, the method \textsc{RepositionContacts} is called.
It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositioning allows to increase the odds that the contact can be maintained at the next step.
Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

\begin{algorithm}
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
		\State State $newState$
		\State $newState.q0 = \mathbf{q}^0$
		\State $newState.freeLimbs = pState.freeLimbs$
		\State /*First try to maintain previous contacts*/
		\State $nbContactsBroken = 0$
		\For {\textbf{each} Limb $k$ in $pState.contactLimbs$}
			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nbContactsBroken += 1$
				\If {$nbContactsBroken > 1$}				
					\State \textbf{return} $NULL$
				\EndIf				
				\State $push(newState.freeLimbs,k)$
			\Else 					
				\State $push(newState.contactLimbs,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} Limb $k$ in $pState.freeLimbs$}
			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
				\State $push(newState.contactLimbs,k)$
				\State $remove(newState.freeLimbs,k)$		
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
			\State \textbf{return} $newState$
		\Else
			\State \textbf{return} $NULL$
		\EndIf
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

@@

The pseudo code for the method \textsc{RepositionContacts} is given by Algorithm~\ref{alg:repo}.

@@

\begin{algorithm}
\caption{Performs contact repositioning for one limb} \label{interpolate}
	\begin{algorithmic}[1]
	\Function{RepositionContacts}{$state$}
		\State $i=0$
		\While {$i<length(states.freeLimbs)$}
			\State Limb $k = pop(states.freeLimbs)$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $i+=1$
				\State $push(states.freeLimbs,k)$		
			\EndIf
		\EndWhile
		\State $i=0$
		\While {$i<length(states.contactLimbs)$}
			\State Limb $k = pop(states.contactLimbs)$
			\State Limb $copy = k$
			\State $i+=1$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $push(newState.contactLimbs,copy)$	
			\EndIf
		\EndWhile
		/*Fails if impossible to relocate any effector*/
		\State \textbf{return} $FAILURE$
	\EndFunction
\end{algorithmic}
\label{alg:repo}
\end{algorithm}

>>>>

% !TEX root =  ../main_tro.tex
\subsubsection{Pseudo-code of the Algorithm}
\label{app:contact}

@@

\begin{algorithm}[!tbp]
\caption{Discretization of a path} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{Interpolate}{$s0$,$\mathbf{Q}^0$, $MAX\_TRIES$}
		\State $list<$State$>$ $states = [s0]$
		\State $nb\_fail = 0$ 
		\State $i = 1;$ /*Current index in the list*/
		\While {$i < length(\mathbf{Q}^0)$}
			\State State $pState = last\_element(states)$
			\State State $s =$ \textsc{GenFullBody}$(pState, \mathbf{Q}^0[i])$
			\If {$s != NULL $}
				\State $nb\_fail = 0$
				\State $i += 1$
				\State \textbf{return} $\mathbf{q}^{0}$
			\Else
				\State $nb\_fail += 1$
				\If {$nb\_fail == MAX\_TRIES$}
					\State \textbf{return} $FAILURE$
				\EndIf				
				\State $s = $\textsc{IntermediateContactState}$(pState)$
			\EndIf
			\State $push\_back(states, s)$
		\EndWhile
		\State \textbf{return} $states$
	\EndFunction
\end{algorithmic}
\label{alg:interpolate}
\end{algorithm}

At each step, \textsc{GenFullBody} is called with the previous state as a parameter, as well
as a new root configuration. \textsc{GenFullBody} returns a new contact configuration, if it succeeded
in computing a configuration with only one contact switch occurring.
Otherwise, the method \textsc{IntermediateContactState} is called.
It repositions one end effector (either a free limb, or the oldest active contact) towards a new contact position if possible.
This repositioning allows to increase the odds that the contact can be maintained at the next step.
Algorithm~\ref{alg:pi} gives the pseudo code for \textsc{GenFullBody}.

\begin{algorithm}[!tbp]
\caption{Full body contact generation method} \label{interpolate}
	\begin{algorithmic}[1]
	%~ \Function{GenerateConfiguration}{}
	\Function{\textsc{GenFullBody}}{$pState$,$\mathbf{q}^0$}
		\State State $newState$
		\State $newState.q0 = \mathbf{q}^0$
		\State $newState.freeLimbs = pState.freeLimbs$
		\State /*First try to maintain previous contacts*/
		\State $nbContactsBroken = 0$
		\For {\textbf{each} Limb $k$ in $pState.contactLimbs$}
			\If {$!$\textsc{MaintainContact}$(pState,\mathbf{q}^0,k)$}
				\State $nbContactsBroken += 1$
				\If {$nbContactsBroken > 1$}				
					\State \textbf{return} $NULL$
				\EndIf				
				\State $push(newState.freeLimbs,k)$
			\Else 					
				\State $push(newState.contactLimbs,k)$
			\EndIf
		\EndFor
		\For {\textbf{each} Limb $k$ in $pState.freeLimbs$}
			\If {\textsc{GenerateContact}$(\mathbf{q}^0,k)$}	
				\State $push(newState.contactLimbs,k)$
				\State $remove(newState.freeLimbs,k)$		
				\State \textbf{return} $newState$
			\EndIf
		\EndFor
		\If {\textsc{IsInStaticEquilibrium}$(newState)$}
			\State \textbf{return} $newState$
		\Else
			\State \textbf{return} $NULL$
		\EndIf
	\EndFunction
\end{algorithmic}
\label{alg:pi}
\end{algorithm}

@@

The pseudo code for the method \textsc{IntermediateContactState} is given by Algorithm~\ref{alg:repo}.

@@

\begin{algorithm}[!tbp]
\caption{Adds or repositions a contact for one limb} \label{interpolate}
	\begin{algorithmic}[1]
	\Function{IntermediateContactState}{$state$}
		\State $i=0$
		\While {$i<length(states.freeLimbs)$}
			\State Limb $k = pop(states.freeLimbs)$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $i+=1$
				\State $push(states.freeLimbs,k)$		
			\EndIf
		\EndWhile
		\State $i=0$
		\While {$i<length(states.contactLimbs)$}
			\State Limb $k = pop(states.contactLimbs)$
			\State Limb $copy = k$
			\State $i+=1$
			\If {\textsc{GenerateContact}$(state.q0,k)$}	
				\State $push(newState.contactLimbs,k)$			
				\State \textbf{return}
			\Else
				\State $push(newState.contactLimbs,copy)$	
			\EndIf
		\EndWhile
		/*Fails if impossible to relocate any effector*/
		\State \textbf{return} $FAILURE$
	\EndFunction
\end{algorithmic}
\label{alg:repo}
\end{algorithm}

### 2017-10-02 body/appendix_hpp.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS, that can be found at \url{http://projects.laas.fr/gepetto/index.php/Software/Main}.
HPP implements the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

@@

Our implementation of the planner is also open source, and can be found at \url{https://github.com/stonneau/hpp-rbprm}.

>>>>

\section{Source code of our planner}
\label{app:hpp}
Our planner is implemented using the Humanoid Path Planner (HPP) software, introduced in~\cite{7759083}.
HPP is an open source motion planning framework developed by the Gepetto team at LAAS-CNRS.
HPP implements the standard tools and algorithms used in motion planning,
such as the Bi-RRT planner from which RB-RRT is derived.

@@

Our implementation of the planner is also open source.
Both HPP and our planner can be simply downloaded and compiled by following the instructions on
\url{https://humanoid-path-planner.github.io/hpp-doc/download.html?branch=rbprm}.

### 2017-10-02 body/appendix_optim.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox}), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

>>>>

% !TEX root =  ../main.tex
\section{Addressing $\mathcal{P}_3$}
\label{app:optim}
The stair climbing and the standing up scenarios were validated with the trajectory optimization scheme provided in \citeauthor{Carpentier2016}. 
%~ \deladp{The optimization is formulated as an optimal control problem that aims at generating a valid trajectory for the center of mass.}
To address the other scenarios, we propose a new implementation, entirely open source (\url{https://github.com/stonneau/python_sandbox/releases/tag/tro_paper} ), which can be integrated directly in our motion planner software. This formulation uses the center of mass acceleration and angular momentum as input variables, while previously the contact forces were used.
The center-of-mass trajectory resulting from the optimization is then turned into a collision-free whole-body trajectory.

### 2017-10-02 body/appendix_roms.tex
corrections fro mrev 1
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

We detail our method to generate the volumes $W$ used
in RB-RRT, with the example of HRP-2.
The kinematic chain is split into four limbs $R^k$.
The arms are connected to the shoulders, and the legs to the root.
The obtained volumes $W$ are shown in Figure~\ref{fig:hrp2_w}.

@@

To generate a volume $W^k$, we proceed as follows:
\begin{enumerate}
\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
\item The resulting polytope can contain a very large number of faces. A last step is thus to simplify it in a conservative way with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
\end{enumerate}

>>>>

We detail our method to generate the volumes $W$ used
in RB-RRT, with the example of HRP-2.
The kinematic tree is split into four limbs $R^k$.
The arms are connected to the shoulders, and the legs to the root.
The obtained volumes $W$ are shown in Figure~\ref{fig:hrp2_w}.

@@

To generate a volume $W^k$, we proceed as follows:
\begin{enumerate}
\item Generate randomly $N$ valid limb configurations for $R^k$, for $N$ really large (say $100000$);
\item For each configuration, store the 3D position of the end effector joint relatively to the root of $R^k$; then compute the convex hull of the resulting point cloud;
\item The resulting polytope can contain a very large number of faces. A last step is thus to simplify it with the blender decimate tool (\url{http://wiki.blender.org/index.php/Doc:2.4/Manual/Modifiers/Generate/Decimate}). This tool removes a user-defined amount of vertices (and faces) of the polytope, thus resulting
in a convervative approximation of the original shape. For HRP-2 we apply the operator with a ratio of $0.06$, resulting in a polytope of 38 faces for the arms and the legs.
\end{enumerate}

### 2017-10-02 body/rbprm.tex
most modifs included, missing pdf of reviewer 1 minor comments, sota, and some others

The strong hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
\begin{equation} \label{eq:creach}
\forall t \in [0,1], \mathbf{q}^0(t) \in C_{Reach}^0 \Rightarrow \mathbf{q}^0(t)  \in C_{Equil}^0
\end{equation}
% for \gls{cluttered} problems.
We call  $C_{Reach}^0$ the \textit{contact reachable workspace}, and detail its construction in the following.
The validity of this hypothesis is discussed in depth in Section~\ref{sec:discussion}.
%~ We describe the construction of $C_{Reach}^0$ in the following.

@@

\subsection{The compromising reachability condition}
\label{sec:scaling}
The ideal shape $B^*, W^0 \subset B* \subset B^{\textrm{\it Suf}}$ that defines a necessary \textbf{and} sufficient condition for contact creation has no explicit definition.
Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.

>>>>

The main hypothesis of this work is that for a large variety of locomotion tasks, we can define a space  $C_{Reach}^0 \simeq C_{Contact}^0$, such that 
\begin{equation} \label{eq:creach}
\forall t \in [0,1], \mathbf{q}^0(t) \in C_{Reach}^0 \Rightarrow \mathbf{q}^0(t)  \in C_{Equil}^0
\end{equation}
% for \gls{cluttered} problems.
We call  $C_{Reach}^0$ the \textit{contact reachable workspace}, and detail its construction in the following.
The validity of this hypothesis is discussed in depth in Section~\ref{sec:discussion}.
%~ We describe the construction of $C_{Reach}^0$ in the following.

@@

\subsection{The compromising reachability condition}
\label{sec:scaling}
The ideal shape $B^*, W^0 \subset B^* \subset B^{\textrm{\it Suf}}$ would define a necessary \textbf{and} sufficient condition for contact creation. 
It would guarantee that any root configuration $\mathbf{q}^{0} \in B^*$ would result in a contact configuration, while any $\mathbf{q}^{0} \notin B^*$ could not.
To our knowledge  $B^*$ has no explicit definition.
Therefore, we approximate $B^*$ to define the contact reachable space $C_{Reach}^0$.
